An Experimental Study of Sorting and Branch Prediction

### 2007-01-29 docs/branch_paper/paper.tex
Make pictures fit better

\category{E.5}{Data}{Files}[Sorting/Searching]
\category{C.1.1}{Computer Systems Organization}{Processor Architectures, Single Data Stream Architectures}[Pipeline architectures]
\terms{Algorithms, Experimentation, Measurement, Performance}
\keywords{Sorting, Branch prediction}

@@

\begin{figure}
\centering
\begin{tabular}{cc}
\includegraphics[width=0.5\textwidth]{plots/selection_insertion_sort_branches.eps} & \includegraphics[width=0.5\textwidth]{plots/bubble_shaker_sort_branches.eps} \\
(a) & (b) \\
\end{tabular}
\caption{(a) Shows the branch mispredictions for selection and insertion sort
using bimodal and two-level adaptive predictors both having 4096 table entries.
It is noteworthy the simple bimodal predictor has essentially identical
performance to the two-level adaptive predictors. (b) Shows the branch
mispredictions for bubble and shaker sort using bimodal and two-level adaptive
predictors, again both having 4096 table entries. For bubble sort the bimodal
and two-level adaptive predictors have close to identical performance. However,
for shaker sort the simple bimodal predictor results in fewer mispredictions
than the two-level adaptive predictor. We also note that as a result of the
partial sorting performed by bubble and shaker sort they incur far more branch
mispredictions per key than other $O(n^2)$ sorting algorithms.
\texttt{sim-bpred} was used to simulate the branch predictors in both (a) and
(b). The two-level adaptive predictors used 10-bit history registers.}
\label{elementary_sorts}
\end{figure}

@@

\begin{figure}
\centering
\begin{tabular}{cc}
\includegraphics[width=0.5\textwidth]{plots/bubble_predictability.eps} & \includegraphics[width=0.5\textwidth]{plots/bubble_sweep_misses.eps} \\
(a) & (b) \\
\end{tabular}
\caption{(a) Shows $CP(Q^k_{avg})$ as the sweep number, $k$, increases (i.e. as
the outer-loop iterates). This gives a good approximation to the predictability
of the comparison branch of bubble sort. (b) Shows the number of correctly
predicted branches in bubble sort, as well as the number of executed branches.
Initially the number of correctly predicted branches is close to the number
executed, however, the number of correctly predicted branches rapidly decreases
as the sweep number increases. Gradually, the branches become predictable again.
These trends correspond to the variation in predictability shown in (a). The
results of (b) were obtained using our own software simulated bimodal predictor
averaged over a large number of arrays with 8192 elements. Note also that in (b)
the version of bubble sort used does not terminate early if no swaps are
performed for an entire execution of the inner-loop.  }
\label{bubble_predictability}
\end{figure}

@@

\begin{figure}
\centering
\begin{tabular}{cc}
\includegraphics[width=0.5\textwidth]{plots/elementary_cache1_misses.eps} & \includegraphics[width=0.5\textwidth]{plots/elementary_cycles.eps}\\
(a) & (b) \\
\end{tabular}
\caption{(a) Shows the level 1 cache misses of the elementary sorting
algorithms.  These results were obtained by using \texttt{sim-cache} to simulate
an 8KB direct mapped level 1 cache with 32 byte cache lines. Note that there are
no level 2 cache misses since the simulated 2nd level cache is 2MB in size.  (b)
Shows the number of cycles per key for the elementary sorting algorithms.
Despite bubble and shaker sort having better cache performance than selection
sort, the vast number of branch mispredictions (see Figure
\ref{elementary_sorts}(b)) they cause make them less efficient than selection
sort in practice.  These results were measured using hardware performance
counters on a Pentium 4.}
\label{shaker_plots}
\end{figure}

@@

\begin{figure}
\centering
\begin{tabular}{cc}
\includegraphics[width=0.5\textwidth]{plots/shellsort_branch_misses.eps} & \includegraphics[width=0.5\textwidth]{plots/shell_heap_cache_misses.eps} \\
(a) & (b) \\
\includegraphics[width=0.5\textwidth]{plots/shell_heap_L2_misses.eps} & \includegraphics[width=0.5\textwidth]{plots/shell_heap_cycles.eps} \\
(a) & (b) \\
\end{tabular}
\caption{(a) Shows the average branches misses per key for shellsort using a
bimodal and two-level adaptive predictor, both having 4096 table entries.  On
this occasion, the two-level adaptive predictor out-performs the simpler bimodal
predictor. \texttt{sim-bpred} was used to simulate these branch predictors.  (b)
and (c) show that our basic heapsort implementation has inferior cache
performance at both levels of the cache compared to our shellsort
implementation. Shellsort incurs only a small number of cache misses because on
average a key moves just 0.9744 places per iteration. These results use
\texttt{sim-cache} to simulate an 8KB direct mapped level 1 cache and a 2MB
direct mapped level 2 cache both with 32 byte cache lines. Finally (d) shows
that our shellsort implementation out-performs our heapsort implementation,
these results were measured using Pentium 4 hardware performance counters.}
\label{shellsort_results_figs}
\end{figure}

@@

\begin{figure}
\centering
\begin{tabular}{cc}
\includegraphics[width=0.5\textwidth]{plots/heapsort_branch_counts.eps} & \includegraphics[width=0.5\textwidth]{plots/heapsort_branch_misses.eps}\\
(a) & (b) \\
\includegraphics[width=0.5\textwidth]{plots/heapsort_cycle_counts.eps} & \includegraphics[width=0.5\textwidth]{plots/heapsort_8heap_branches.eps} \\
(c) & (d) \\
\end{tabular}
\caption{Branch prediction and performance characteristics of our heapsort
implementations. (a) Shows the average number of branches per key measured using
\texttt{sim-bpred}. (b) Shows the average branch mispredictions per key for
tables of 4096 predictors, again measured using \texttt{sim-bpred}. The
two-level adaptive (which used a 10-bit history register) and bimodal predictors
have close to identical performance, and to avoid cluttering the diagram only
one instance of a two-level adaptive predictor is shown, the remainder of the
predictors are bimodal. (c) Shows the cycle counts of the algorithms measured
using Pentium 4 hardware performance counters. Note that the Pentium 4 has 32
byte cache lines in its first level cache, and 64 byte cache lines in its second
level cache.  Finally (d) shows the predictability of the unrolled branches
associated with an 8-heap, measured with our own software simulated bimodal
predictor.}
\label{heapsort_results}
\end{figure}

@@

\begin{figure}
\centering
\begin{tabular}{cc}
\includegraphics[width=0.5\textwidth]{plots/mergesort_branches_per_key.eps} & \includegraphics[width=0.5\textwidth]{plots/mergesort_branch_misses.eps} \\
(a) & (b) \\
\includegraphics[width=0.5\textwidth]{plots/mergesort_level_2_misses.eps} & \includegraphics[width=0.5\textwidth]{plots/mergesort_cycle_counts.eps} \\
(c) & (d) \\
\end{tabular}
\caption{(a) Shows the branches per key executed by our mergesort variations.
The sudden increase in the number of executed branches for multi-mergesort
results from the use of the multi-way merge when the data no longer fits within
the cache. (b) Shows the number of mispredictions per key, all predictors had
4096 table entries. Bimodal predictors were used for base and tiled mergesort.
These results were generated using \texttt{sim-bpred}, the two-level adaptive
predictor had a 10-bit history register.  (c) Shows the level 2 cache misses per
key, base mergesort has by far the worst cache behaviour, while multi-mergesort
has slightly better cache performance than tiled mergesort. These results were
generated by using \texttt{sim-cache} to simulate a direct mapped 2MB cache with
32-byte cache lines. Finally (d) shows the cycles per key of our algorithms
measured using Pentium 4 hardware performance counters. Despite
multi-mergesort's superior cache performance, its heightened instruction count
and branch mispredictions mean that the less cache conscious tiled mergesort
performs the best out of the algorithms.}
\label{mergesort_branch_results}
\end{figure}

@@

\begin{figure}
\centering
\begin{tabular}{cc}
\includegraphics[width=0.5\textwidth]{plots/counter_base_quicksort__median_of_3__0.eps} & \includegraphics[width=0.5\textwidth]{plots/counter_memory_tuned_quicksort_0.eps} \\
(a) Basic quicksort & (b) Memory-tuned quicksort \\ 
\includegraphics[width=0.5\textwidth]{plots/counter_multi_quicksort__binary_search__0.eps} & \includegraphics[width=0.5\textwidth]{plots/counter_multi_quicksort__sequential_search__0.eps} \\
(c) Multi-quicksort (binary search) & (d) Multi-quicksort (sequential search)
\end{tabular}
\caption{Overview of branch prediction behaviour in our quicksort
implementations. Every figure shows the behaviour of the $i$ and $j$ branches
when using a median-of-3 pivot. As described in Section
\ref{quicksort_branch_results_text}, these branches are about 60\% biased and
64\% predictable when using the median-of-3. In (a) the \texttt{median} branch
is the combined results of the branches which compute the median-of-3 (these
branches are also executed for (b), (c) and (d)).  Comparing (a) with (b), (c)
and (d), we see that the \texttt{insertion} branch associated with its insertion
sort is slightly less predictable than in the other variations. This is due to
it running as a post-pass. Finally, comparing (c) with (d) we see that the
binary search branches of (c), \texttt{binary left} and \texttt{binary right},
are very unpredictable compared to the \texttt{sequential} branch of (d).  }
\label{branch_prediction_quicksort}
\end{figure}

@@

\begin{figure}
\centering
\begin{tabular}{cc}
\includegraphics[width=0.5\textwidth]{plots/quicksort_cache_misses.eps} & \includegraphics[width=0.5\textwidth]{plots/quicksort_instruction_counts.eps} \\
(a) & (b) \\ 
\includegraphics[width=0.5\textwidth]{plots/quicksort_branch_misses.eps} & \includegraphics[width=0.5\textwidth]{plots/quicksort_cycles.eps} \\
(c) & (d) \\
\end{tabular}
\caption{(a) Shows the 2nd level cache misses for our quicksort implementations.
As is expected, the multi-quicksort implementations have the best cache
performance. These results were generated using \texttt{sim-cache} to simulate a
2MB direct-mapped cache with 32-byte cache lines.  (b) Shows the instruction
counts of our quicksort implementations, when the data sets no longer fit within
the cache the multi-quicksort implementations show a large increase in
instruction count, these results were also generated using SimpleScalar. (c)
Shows the simulated branch mispredictions per key for our quicksort
implementations. All predictors use 4096 table entries and are bimodal
predictors except where indicated. Multi-quicksort using binary search is the
only implementation which gives varying results from the use of a two-level
adaptive predictor. These results were generated using \texttt{sim-bpred}. (d)
Shows the cycles per key for our quicksort implementations measured using
Pentium 4 hardware performance counters.  }
\label{quicksort_plots}
\end{figure}

@@

\begin{figure}
\centering
\begin{tabular}{cc}
\includegraphics[width=0.5\textwidth]{plots/radix_cycles.eps} & \includegraphics[width=0.5\textwidth]{plots/radix_versus_quick_level_2_misses.eps} \\
(a) & (b) \\
\includegraphics[width=0.5\textwidth]{plots/radix_branch_misses.eps} & \includegraphics[width=0.5\textwidth]{plots/radix_branches.eps} \\
(c) & (d) \\
\end{tabular}
\caption{(a) Shows the cycles per key of our radix sort implementations
alongside our base quicksort implementation, measured using Pentium 4 hardware
performance counters. (b) Shows the level 2 cache misses incurred by our radix
sort implementations and a multi-quicksort implementation for a 2MB direct
mapped cache with 32-byte cache lines. These results were generated using
\texttt{sim-cache}.  (c) Shows the average branch mispredictions per key
incurred by our radix sort implementations while using bimodal predictors with
4096 table entries, the results for two-level adaptive predictors are extremely
similar. Finally, (d) shows the average number of branches per key executed by
our  implementations. The results of (c) and (d) were generated using
\texttt{sim-bpred}.}
\label{radixsort_results_pictures}
\end{figure}

>>>>

\category{E.5}{Data}{Files}[Sorting/Searching]
\category{C.1.1}{Computer Systems Organization}{Processor Architectures, Other
Architecture Styles}[Pipeline processors]
\terms{Algorithms, Experimentation, Measurement, Performance}
\keywords{Sorting, Branch Prediction, Pipeline Architectures, Caching}

@@

\begin{figure}
\centering
\begin{tabular}{cc}
\includegraphics[width=0.48\textwidth]{plots/selection_insertion_sort_branches.eps} & \includegraphics[width=0.48\textwidth]{plots/bubble_shaker_sort_branches.eps} \\
(a) & (b) \\
\end{tabular}
\caption{(a) Shows the branch mispredictions for selection and insertion sort
using bimodal and two-level adaptive predictors both having 4096 table entries.
It is noteworthy the simple bimodal predictor has essentially identical
performance to the two-level adaptive predictors. (b) Shows the branch
mispredictions for bubble and shaker sort using bimodal and two-level adaptive
predictors, again both having 4096 table entries. For bubble sort the bimodal
and two-level adaptive predictors have close to identical performance. However,
for shaker sort the simple bimodal predictor results in fewer mispredictions
than the two-level adaptive predictor. We also note that as a result of the
partial sorting performed by bubble and shaker sort they incur far more branch
mispredictions per key than other $O(n^2)$ sorting algorithms.
\texttt{sim-bpred} was used to simulate the branch predictors in both (a) and
(b). The two-level adaptive predictors used 10-bit history registers.}
\label{elementary_sorts}
\end{figure}

@@

\begin{figure}
\centering
\begin{tabular}{cc}
\includegraphics[width=0.48\textwidth]{plots/bubble_predictability.eps} & \includegraphics[width=0.48\textwidth]{plots/bubble_sweep_misses.eps} \\
(a) & (b) \\
\end{tabular}
\caption{(a) Shows $CP(Q^k_{avg})$ as the sweep number, $k$, increases (i.e. as
the outer-loop iterates). This gives a good approximation to the predictability
of the comparison branch of bubble sort. (b) Shows the number of correctly
predicted branches in bubble sort, as well as the number of executed branches.
Initially the number of correctly predicted branches is close to the number
executed, however, the number of correctly predicted branches rapidly decreases
as the sweep number increases. Gradually, the branches become predictable again.
These trends correspond to the variation in predictability shown in (a). The
results of (b) were obtained using our own software simulated bimodal predictor
averaged over a large number of arrays with 8192 elements. Note also that in (b)
the version of bubble sort used does not terminate early if no swaps are
performed for an entire execution of the inner-loop.  }
\label{bubble_predictability}
\end{figure}

@@

\begin{figure}
\centering
\begin{tabular}{cc}
\includegraphics[width=0.48\textwidth]{plots/elementary_cache1_misses.eps} & \includegraphics[width=0.48\textwidth]{plots/elementary_cycles.eps}\\
(a) & (b) \\
\end{tabular}
\caption{(a) Shows the level 1 cache misses of the elementary sorting
algorithms.  These results were obtained by using \texttt{sim-cache} to simulate
an 8KB direct mapped level 1 cache with 32 byte cache lines. Note that there are
no level 2 cache misses since the simulated 2nd level cache is 2MB in size.  (b)
Shows the number of cycles per key for the elementary sorting algorithms.
Despite bubble and shaker sort having better cache performance than selection
sort, the vast number of branch mispredictions (see Figure
\ref{elementary_sorts}(b)) they cause make them less efficient than selection
sort in practice.  These results were measured using hardware performance
counters on a Pentium 4.}
\label{shaker_plots}
\end{figure}

@@

\begin{figure}
\centering
\begin{tabular}{cc}
\includegraphics[width=0.48\textwidth]{plots/shellsort_branch_misses.eps} & \includegraphics[width=0.48\textwidth]{plots/shell_heap_cache_misses.eps} \\
(a) & (b) \\
\includegraphics[width=0.48\textwidth]{plots/shell_heap_L2_misses.eps} & \includegraphics[width=0.48\textwidth]{plots/shell_heap_cycles.eps} \\
(a) & (b) \\
\end{tabular}
\caption{(a) Shows the average branches misses per key for shellsort using a
bimodal and two-level adaptive predictor, both having 4096 table entries.  On
this occasion, the two-level adaptive predictor out-performs the simpler bimodal
predictor. \texttt{sim-bpred} was used to simulate these branch predictors.  (b)
and (c) show that our basic heapsort implementation has inferior cache
performance at both levels of the cache compared to our shellsort
implementation. Shellsort incurs only a small number of cache misses because on
average a key moves just 0.9744 places per iteration. These results use
\texttt{sim-cache} to simulate an 8KB direct mapped level 1 cache and a 2MB
direct mapped level 2 cache both with 32 byte cache lines. Finally (d) shows
that our shellsort implementation out-performs our heapsort implementation,
these results were measured using Pentium 4 hardware performance counters.}
\label{shellsort_results_figs}
\end{figure}

@@

\begin{figure}
\centering
\begin{tabular}{cc}
\includegraphics[width=0.48\textwidth]{plots/heapsort_branch_counts.eps} & \includegraphics[width=0.48\textwidth]{plots/heapsort_branch_misses.eps}\\
(a) & (b) \\
\includegraphics[width=0.48\textwidth]{plots/heapsort_cycle_counts.eps} & \includegraphics[width=0.48\textwidth]{plots/heapsort_8heap_branches.eps} \\
(c) & (d) \\
\end{tabular}
\caption{Branch prediction and performance characteristics of our heapsort
implementations. (a) Shows the average number of branches per key measured using
\texttt{sim-bpred}. (b) Shows the average branch mispredictions per key for
tables of 4096 predictors, again measured using \texttt{sim-bpred}. The
two-level adaptive (which used a 10-bit history register) and bimodal predictors
have close to identical performance, and to avoid cluttering the diagram only
one instance of a two-level adaptive predictor is shown, the remainder of the
predictors are bimodal. (c) Shows the cycle counts of the algorithms measured
using Pentium 4 hardware performance counters. Note that the Pentium 4 has 32
byte cache lines in its first level cache, and 64 byte cache lines in its second
level cache.  Finally (d) shows the predictability of the unrolled branches
associated with an 8-heap, measured with our own software simulated bimodal
predictor.}
\label{heapsort_results}
\end{figure}

@@

\begin{figure}
\centering
\begin{tabular}{cc}
\includegraphics[width=0.48\textwidth]{plots/mergesort_branches_per_key.eps} & \includegraphics[width=0.48\textwidth]{plots/mergesort_branch_misses.eps} \\
(a) & (b) \\
\includegraphics[width=0.48\textwidth]{plots/mergesort_level_2_misses.eps} & \includegraphics[width=0.48\textwidth]{plots/mergesort_cycle_counts.eps} \\
(c) & (d) \\
\end{tabular}
\caption{(a) Shows the branches per key executed by our mergesort variations.
The sudden increase in the number of executed branches for multi-mergesort
results from the use of the multi-way merge when the data no longer fits within
the cache. (b) Shows the number of mispredictions per key, all predictors had
4096 table entries. Bimodal predictors were used for base and tiled mergesort.
These results were generated using \texttt{sim-bpred}, the two-level adaptive
predictor had a 10-bit history register.  (c) Shows the level 2 cache misses per
key, base mergesort has by far the worst cache behaviour, while multi-mergesort
has slightly better cache performance than tiled mergesort. These results were
generated by using \texttt{sim-cache} to simulate a direct mapped 2MB cache with
32-byte cache lines. Finally (d) shows the cycles per key of our algorithms
measured using Pentium 4 hardware performance counters. Despite
multi-mergesort's superior cache performance, its heightened instruction count
and branch mispredictions mean that the less cache conscious tiled mergesort
performs the best out of the algorithms.}
\label{mergesort_branch_results}
\end{figure}

@@

\begin{figure}
\centering
\begin{tabular}{cc}
\includegraphics[width=0.48\textwidth]{plots/counter_base_quicksort__median_of_3__0.eps} & \includegraphics[width=0.48\textwidth]{plots/counter_memory_tuned_quicksort_0.eps} \\
(a) Basic quicksort & (b) Memory-tuned quicksort \\ 
\includegraphics[width=0.48\textwidth]{plots/counter_multi_quicksort__binary_search__0.eps} & \includegraphics[width=0.48\textwidth]{plots/counter_multi_quicksort__sequential_search__0.eps} \\
(c) Multi-quicksort (binary search) & (d) Multi-quicksort (sequential search)
\end{tabular}
\caption{Overview of branch prediction behaviour in our quicksort
implementations. Every figure shows the behaviour of the $i$ and $j$ branches
when using a median-of-3 pivot. As described in Section
\ref{quicksort_branch_results_text}, these branches are about 60\% biased and
64\% predictable when using the median-of-3. In (a) the \texttt{median} branch
is the combined results of the branches which compute the median-of-3 (these
branches are also executed for (b), (c) and (d)).  Comparing (a) with (b), (c)
and (d), we see that the \texttt{insertion} branch associated with its insertion
sort is slightly less predictable than in the other variations. This is due to
it running as a post-pass. Finally, comparing (c) with (d) we see that the
binary search branches of (c), \texttt{binary left} and \texttt{binary right},
are very unpredictable compared to the \texttt{sequential} branch of (d).  }
\label{branch_prediction_quicksort}
\end{figure}

@@

\begin{figure}
\centering
\begin{tabular}{cc}
\includegraphics[width=0.48\textwidth]{plots/quicksort_cache_misses.eps} & \includegraphics[width=0.48\textwidth]{plots/quicksort_instruction_counts.eps} \\
(a) & (b) \\ 
\includegraphics[width=0.48\textwidth]{plots/quicksort_branch_misses.eps} & \includegraphics[width=0.48\textwidth]{plots/quicksort_cycles.eps} \\
(c) & (d) \\
\end{tabular}
\caption{(a) Shows the 2nd level cache misses for our quicksort implementations.
As is expected, the multi-quicksort implementations have the best cache
performance. These results were generated using \texttt{sim-cache} to simulate a
2MB direct-mapped cache with 32-byte cache lines.  (b) Shows the instruction
counts of our quicksort implementations, when the data sets no longer fit within
the cache the multi-quicksort implementations show a large increase in
instruction count, these results were also generated using SimpleScalar. (c)
Shows the simulated branch mispredictions per key for our quicksort
implementations. All predictors use 4096 table entries and are bimodal
predictors except where indicated. Multi-quicksort using binary search is the
only implementation which gives varying results from the use of a two-level
adaptive predictor. These results were generated using \texttt{sim-bpred}. (d)
Shows the cycles per key for our quicksort implementations measured using
Pentium 4 hardware performance counters.  }
\label{quicksort_plots}
\end{figure}

@@

\begin{figure}
\centering
\begin{tabular}{cc}
\includegraphics[width=0.48\textwidth]{plots/radix_cycles.eps} & \includegraphics[width=0.48\textwidth]{plots/radix_versus_quick_level_2_misses.eps} \\
(a) & (b) \\
\includegraphics[width=0.48\textwidth]{plots/radix_branch_misses.eps} & \includegraphics[width=0.48\textwidth]{plots/radix_branches.eps} \\
(c) & (d) \\
\end{tabular}
\caption{(a) Shows the cycles per key of our radix sort implementations
alongside our base quicksort implementation, measured using Pentium 4 hardware
performance counters. (b) Shows the level 2 cache misses incurred by our radix
sort implementations and a multi-quicksort implementation for a 2MB direct
mapped cache with 32-byte cache lines. These results were generated using
\texttt{sim-cache}.  (c) Shows the average branch mispredictions per key
incurred by our radix sort implementations while using bimodal predictors with
4096 table entries, the results for two-level adaptive predictors are extremely
similar. Finally, (d) shows the average number of branches per key executed by
our  implementations. The results of (c) and (d) were generated using
\texttt{sim-bpred}.}
\label{radixsort_results_pictures}
\end{figure}
